---
title: "Survival analysis of TCGA data based on RNA-Seq-derived immunophenotype"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Refs:

- https://stats.stackexchange.com/questions/29135/difference-between-survival-analysis-and-classification
- https://cran.r-project.org/web/packages/party/party.pdf / ctree
- For a general description of the methodology see Hothorn, Hornik and Zeileis (2006) and Hothorn, Hornik, van de Wiel and Zeileis (2006). Introductions for novices can be found in Strobl et al. (2009) and at http://github.com/christophM/overview-ctrees.git. 

Torsten Hothorn, Kurt Hornik, Mark A. van de Wiel and Achim Zeileis (2006). A Lego System for Conditional Inference. The American Statistician, 60(3), 257–263.
Torsten Hothorn, Kurt Hornik and Achim Zeileis (2006). Unbiased Recursive Partitioning: A Conditional Inference Framework. Journal of Computational and Graphical Statistics, 15(3), 651–674. Preprint available from http://statmath.wu-wien.ac.at/~zeileis/papers/Hothorn+Hornik+Zeileis-2006.pdf

Load data

```{r}
library(party)
library(survival)
library(dplyr)
library(ggplot2)
library(memisc)
library(reshape2)

df = read.table("Tree.txt", header = T, sep = "\t") %>% filter(!is.na(OS.uncorrected))

df$Stage2 = with(df, cases(
  Stage == "i" -> 1,
  Stage == "ii" -> 2,
  Stage == "iii" -> 3,
  Stage == "iv" -> 4,
  T -> NA
))

df$Tumor.site2 = with(df, cases(
  Tumor.site == "Distant Metastasis" -> "D.M.",
  Tumor.site == "Primary Tumor" -> "P.T.",
  Tumor.site == "Regional Cutaneous o" -> "R.C.",
  Tumor.site == "Regional Lymph Node" -> "R.L.N.",
  T -> NA
))

df$Tumor.site2 = as.factor(df$Tumor.site2)

df$Stage3 = as.factor(with(df, ifelse(Stage == "" | Stage == "0", NA, as.character(Stage))))
```

```{r}
ggplot(df %>% dplyr::select(IGH.clonality, 
                     IGH.coverage.by.MiXCR,
                     Ratio.IGG1.to.IGH.by.MiXCR,
                     TRB.coverage.by.MiXCR,
                     TRA.coverage.by.MiXCR) %>% melt,
       aes(x=value, color=variable)) +
  geom_density() +
  facet_wrap(~variable, scales = "free") +
  theme_bw()
```

Decision tree learning parameters - we don't correct P-values for multiple testing; minimal size of bucket is 30 donors

```{r}
ctree_opts = ctree_control(testtype = "Univariate", #mincriterion = 0.8, #
                           minbucket = 45)

make_tree = function(formula) {
  dfct = ctree(formula, data = df, controls = ctree_opts)
  plot(dfct)
}
```

Stage correction for OS works well

```{r}
make_tree(Surv(OS.uncorrected, Dead) ~ Stage3)
make_tree(Surv(OS.corrected, Dead) ~ Stage3)
```

IG-based model plus tumor site (has no effect)

```{r}
make_tree(Surv(OS.corrected, Dead) ~ Tumor.site2 + 
            IGH.clonality + 
            IGH.coverage.by.MiXCR + 
            Ratio.IGG1.to.IGH.by.MiXCR)
```

Adding TRB coverage and clonality

```{r fig.width=16, fig.height=8}
make_tree(Surv(OS.corrected, Dead) ~ 
            IGH.clonality + 
            IGH.coverage.by.MiXCR + 
            Ratio.IGG1.to.IGH.by.MiXCR +
            TRB.coverage.by.MiXCR +
            TRB.clonality)
```

Lets look who performs better among Clusters of Differentiation

```{r}
make_tree(Surv(OS.corrected, Dead) ~  
                Tumor.site2 + 
                CD3.coverage.by.kallisto + 
                CD8.coverage.by.kallisto +
                Ratio.CD138.to.CD20.by.kallisto +
                Ratio.CD138.to.CD45.by.kallisto +
                Ratio.CD319.to.CD45.by.kallisto)
```

Lets put everything together

```{r fig.width=16, fig.height=8}
make_tree(Surv(OS.corrected, Dead) ~ 
                Tumor.site2 + 
                CD3.coverage.by.kallisto + 
                CD8.coverage.by.kallisto +
                Ratio.CD138.to.CD20.by.kallisto +
                Ratio.CD138.to.CD45.by.kallisto +
                Ratio.CD319.to.CD45.by.kallisto +
                IGH.clonality + 
                IGH.coverage.by.MiXCR + 
                Ratio.IGG1.to.IGH.by.MiXCR +
                TRB.coverage.by.MiXCR +
                TRB.clonality)
```